Precise temporal localisation of M/EEG effects with Bayesian generalised additive multilevel models

### 2025-04-02 brms_meeg.qmd
minor updates to cache

---
title: "Precise temporal localisation of M/EEG effects with Bayesian generalised additive multilevel models"
shorttitle: "Bayesian modelling of M/EEG data"
author:
  - name: Ladislas Nalborczyk
    corresponding: true
    orcid: 0000-0002-7419-9855
    email: ladislas.nalborczyk@cnrs.fr
    url: https://lnalborczyk.github.io
    affiliations:
      - id: id1
        name: "Aix Marseille Univ, CNRS, LPL"
        address: "5 avenue Pasteur"
        city: "13100 Aix-en-Provence"
        country: France
  - name: Paul Bürkner
    orcid: 0000-0001-5765-8995
    url: https://paulbuerkner.com
    affiliations:
      - name: "TU Dortmund University, Department of Statistics"
author-note:
  disclosures:
    conflict-of-interest: "The authors have no conflicts of interest to disclose."
    # financial-support: "This study was supported by Grant [Grant Number] from [Funding Source]."
abstract: "Time-resolved electrophysiological measurements such as those offered by magneto- or electro-encephalography (M/EEG) provide a unique window onto neural activity underlying cognitive processes. Typically, researchers are interested in testing whether and when such measures differ across conditions and/or groups. The conventional approach consists in conducting mass-univariate statistics through time followed by some form of multiplicity correction (e.g., FDR, FWER) or cluster-based inference. However, while allowing efficient error-rates control, cluster-based methods have an important downside: they shift the focus of inference from the timepoint to the cluster level, thus preventing any conclusion to be made about the precise onset or offset of effects (e.g., differences across conditions or groups). Here, we introduce a *model-based* approch for analysing M/EEG timeseriesn such as ERPs or decoding performance through time, and their differences across conditions or groups. This approach relies on Bayesian generalised additive multilevel models and outputs the posterior probabilility of the effect being above 0 (or above chance) at every timestep, while naturally taking into account the temporal dependencies and between-subject variability present in such data. Using both simulated and actual M/EEG data, we show that the proposed approach largely outperforms conventional methods in estimating the onset and offset of M/EEG effects, producing more precise and more reliable estimates. We provide an R package implementing the approach and illustrate how to integrate it into M/EEG statistical pipelines in MNE-Python."
keywords: [EEG, MEG, generalised additive models, mixed-effects models, multilevel models, Bayesian statistics, brms]
floatsintext: true
numbered-lines: true
bibliography:
  - meeg_modelling.bib
  - grateful-refs.bib
suppress-title-page: false
link-citations: true
# see https://www.overleaf.com/learn/latex/Font_typefaces
# and systemfonts::system_fonts()
# pdf-engine: xelatex
# monofont: Roboto
mainfont: CMU Serif
fontsize: 10pt
draftfirst: false
draftall: false
a4paper: true
donotrepeattitle: true
lang: en
toc: true
language:
  citation-last-author-separator: "&"
  email: "email"
  title-block-author-note: "Author note"
  title-block-correspondence-note: "Correspondence concerning this article should be addressed to"
  title-block-role-introduction: "Author roles were classified using the Contributor Role Taxonomy (CRediT; https://credit.niso.org/) as follows:"
format:
  # apaquarto-html: default
  apaquarto-pdf:
    latex-engine: xelatex
    documentmode: doc # man, jou, doc, stu
    keep-tex: true
    include-in-header:
      - text: |
          \usepackage{mathtools}
---

```{r setup, include = FALSE}
# loading packages
library(changepoint)
library(tidyverse)
library(tidybayes)
library(patchwork)
library(MetBrewer)
library(grateful)
library(ggrepel)
library(scales)
library(pakret)
library(scico)
library(knitr)
library(brms)

@@

# setting a default ggplot2 theme
theme_set(theme_light(base_size = 10, base_family = "Open Sans") )

@@

@@

Description of cluster-based approaches [see @sassenhagen2019]... As pointed by the original authors themselves [@maris2007], "there There is a conflict between this interest in localized effects and our choice for a global null hypothesis: by controlling the FA \[false alarm\] rate under this global null hypothesis, one cannot quantify the uncertainty in the spatiotemporal localization of the effect" [@maris2007; @sassenhagen2019]... In contrast, the proposed approach, based on Bayesian generalised additive multilevel models, allows quantifying the probability of effects at the level of timesteps, voxels, sensors (etc), while naturally taking into account spatiotemporal dependencies.

@@

where $\beta_{jk}$ is the weight (coefficient) associated with the $k$th basis function $b_{jk}()$ evaluated at the covariate value $x_{ij}$ for the $j$th smooth function $f_{j}$. Splines coefficients are penalised (usually through the squared of the smooth functions' second derivative) in a way that can be interpreted as a prior on the "wiggliness" of the function...

Terminology: *splines* are functions composed of simpler functions. These simpler functions are basis functions (e.g., a cubic polynomial) and the set of basis functions is a *basis*. Each basis function gets its coefficient and the resultant spline is the sum of these weighted basis functions.

@@

Instead of averaring, obtain the smooth ERP signal from multilevel GAM... less susceptible to outliers (Meulman et al., 2023)...

## Gaussian process regression

A Gaussian process (GP) is a stochastic process that defines the distribution over a collection of random variables indexed by a continuous variable, that is $\{f(t): t \in \mathcal{T}\}$ for some index set $\mathcal{T}$ [@riutort-mayol_practical_2023; @rasmussen2005]. Whereas Bayesian linear regression outputs a distribution over the parameters of some predefind parametric model, the GP approach, in contrast, is a non-parametric approach, in that it finds a distribution over the possible functions that are consistent with the observed data. However, note that nonparametric does not mean there aren't parameters, it means that there are infinitely many parameters.

See also <https://jtimonen.github.io/lgpr-usage/articles/math.html>...

@@

We then fitted a Bayesian GAM using the `brms` package [@brms2017; @brms2018; @nalborczyk2019]. We used the default priors in `brms` (i.e., weakly informative priors). We ran eight Markov Chain Monte-Carlo (MCMC) to approximate the posterior distribution, including each 5000 iterations and a warmup of 2000 iterations, yielding a total of $8 \times (5000-2000) = 24000$ posterior samples to use for inference. Posterior convergence was assessed examining trace plots as well as the Gelman–Rubin statistic $\hat{R}$. The `brms` package uses the same syntax as `r pkrt("mgcv")` for specifying smooth effects. @fig-post-prob-test shows the predictions of this model together with the raw data.

@@

# fitting the GAM
gam <- brm(
    # cubic regression splines with k-1 basis functions
    eeg ~ condition + s(time, bs = "cr", k = 20, by = condition),
    data = ppt_df,
    family = gaussian(),
    warmup = 2000,
    iter = 5000,
    chains = 8,
    cores = 8,
    file = "models/gam.rds"
    )
```

@@

```{r gp, echo = FALSE, eval = FALSE, message = FALSE}
# Now we fit the GP regression model...
# computing eeg_diff and averaging across participants
# group_df <- raw_df %>%
#     group_by(participant, time) %>%
#     pivot_wider(names_from = condition, values_from = eeg, values_fn = mean) %>%
#     mutate(eeg_diff = cond2 - cond1) %>%
#     summarise(eeg_diff = mean(eeg_diff) ) %>%
#     ungroup()

gp_model <- brm(
    # k refers to the number of basis functions for
    # computing Hilbert-space approximate GPs
    # if k = NA (default), exact GPs are computed
    # eeg ~ gp(time, by = condition),
    eeg ~ condition + gp(time, by = condition, k = 20, cov = "exp_quad"),
    data = ppt_df,
    family = gaussian(),
    # prior = gp_priors,
    control = list(adapt_delta = 0.99, max_treedepth = 15),
    backend = "cmdstanr",
    iter = 2000,
    chains = 4,
    cores = 4,
    file = "models/gp.rds"
    )
```

@@

We then compute the posterior probability of the slope for `condition` being above $0+\epsilon$ (@fig-post-prob-test), with $\epsilon := 0.05$, which can be interpreted as the smallest effect size of interest.

@@

# using the function with the two models
plot_post_test(model = gam, eps = 0.1)
```

We can also express this as the ratio of posterior probabilities (i.e., $p/(1-p)$) and visualise the timecourse of this ratio superimposed with the conventional thresholds on evidence ratios (@fig-post-prob-ratio).

Maybe better to use the p-MAP (i.e., density ratio between the null and the MAP)? And/or look at the derivative (e.g., <https://fromthebottomoftheheap.net/2014/05/15/identifying-periods-of-change-with-gams/>)?

```{r fig-post-prob-ratio, echo = FALSE, out.width = "75%", fig.asp = 0.75, fig.cap = "Ratio of posterior probability according to the GAM (on a log10 scale). Timesteps above threshold (10) are highlighted in green."}
# defining some parameters
n_approx <- 1e3
eps <- 0.05
threshold <- 10

@@

# computing the posterior probability that y is above 0
n_posterior_samples <- 24000
prob_y_above_0 <- time_seq %>%
    add_epred_draws(object = gam) %>%
    data.frame() %>%
    dplyr::select(time, condition, .epred, .draw) %>%
    pivot_wider(names_from = condition, values_from = .epred) %>%
    mutate(epred_diff = cond2 - cond1) %>%
    group_by(time) %>%
    summarise(m = mean(epred_diff > 0 + eps) ) %>%
    ungroup() %>%
    mutate(prob_ratio = m / (1 - m) ) %>%
    # replacing infinite values by maximum possible values (number of posterior samples)
    mutate(prob_ratio = ifelse(is.infinite(prob_ratio), n_posterior_samples, prob_ratio) ) %>%
    mutate(prob_ratio = ifelse(prob_ratio == 0, 1/n_posterior_samples, prob_ratio) )

@@

# plotting it
gam_ratio_plot <- prob_y_above_0 %>%
    mutate(above_thres = ifelse(prob_ratio > threshold, 1, NA) ) %>%
    ggplot(aes(x = time, y = prob_ratio) ) +
    geom_raster(
        data = bg_data,
        aes(x = time, y = prob_ratio, fill = fill_value),
        alpha = 0.5,
        interpolate = FALSE,
        inherit.aes = FALSE,
        show.legend = FALSE
        ) +
    scale_fill_scico(palette = "vik", midpoint = log10(1) ) +
    geom_hline(yintercept = 1, linetype = "dashed") +
    geom_hline(yintercept = 3, linetype = "dashed", color = "darkred") +
    geom_hline(yintercept = 1/3, linetype = "dashed", color = "darkred") +
    geom_hline(yintercept = 10, linetype = "dashed", color = "red") +
    geom_hline(yintercept = 1/10, linetype = "dashed", color = "red") +
    geom_hline(yintercept = 100, linetype = "dashed", color = "orangered") +
    geom_hline(yintercept = 1/100, linetype = "dashed", color = "orangered") +
    geom_vline(xintercept = true_onset, linetype = 2) +
    geom_vline(xintercept = true_offset, linetype = 2) +
    geom_line(linewidth = 0.8) +
    geom_point(
        data = . %>% dplyr::filter(!is.na(above_thres) ),
        aes(y = threshold),
        colour = "darkgreen",
        shape = 16,
        size = 3,
        na.rm = TRUE
        ) +
    scale_y_log10(
        labels = label_log(digits = 2),
        breaks = c(1/1e4, 1/1e3, 1/1e2, 1/10, 1, 10, 1e2, 1e3, 1e4)
        # limits = c(0.0001, 1e4)
        ) +
    coord_cartesian(expand = FALSE) +
    labs(
        x = "Time (s)",
        y = expression(Pr(beta>0~"|"~data) / (1 - Pr(beta>0~"|"~data) ) )
        )

@@

# or plotting only the GAM
gam_ratio_plot

# printing the results
# bind_rows(exceeding_times_gam, exceeding_times_gp) %>%
#     mutate(across(.cols = where(is.numeric), .fns = ~round(.x, 2) ) ) %>%
#     relocate(model, .before = cluster_onset)

```

@@

Next we fit a hierarchical/multilevel GAM using summary statistics of ERPs (mean and SD) at the participant level (similar to what is done in meta-analysis). Although it is possible to fit a GAMM at the single-trial level, we present a resource-efficient version of the model that is fitted directly on the by-participant summary statistics (mean and SD)...

@@

We then computed the difference between the true and estimated onset/offset of the ERP difference ($\text{error}:=|\hat{\theta}-\theta|$), according to various `eps` and `threshold` values. Remember that the signal is generated from a truncated Gaussian with an objective onset at 160 ms, a maximum at 250 ms, and an offset at 342 ms. @fig-onset-error shows that the hierarchical GAM can *exactly* recover the true onset and offset values, given some reasonable choice of `eps` and `threshold` values (e.g., a threshold of 20).

```{r fig-onset-error, echo = FALSE, dpi = 300, fig.width = 8, fig.asp = 0.5, fig.cap = "Error function of onset (left) and offset (right) estimation according to various eps and threshold values (according to the hierarchical GAM). Minimum error values are indicated by red crosses."}
# defining values of eps and threshold to test
eps_values <- seq(from = 0, to = 0.1, by = 0.01)
threshold_values <- seq(from = 1, to = 30, by = 1)

# using as many values of time as in the original data
# n_approx <- n_distinct(ppt_df$time)

# number of posterior samples to use
n_post_samples <- 1e3

# initialising results dataframe
results <- crossing(eps = eps_values, threshold = threshold_values) %>%
    mutate(
        estimated_onset = NA, estimated_peak = NA, estimated_offset = NA,
        error_onset = NA, error_peak = NA, error_offset = NA
        )

# which model should we use?
error_model <- meta_gam

# generating a grid of time values to compute predictions for
# time_seq <- crossing(
#     condition = c("cond1", "cond2"),
#     time = seq(min(ppt_df$time), max(ppt_df$time), length.out = n_approx*2-1),
#     ) %>%
#     arrange(condition)

# eps <- 0.05
# preds <- meta_gam$data %>%
#     add_epred_draws(meta_gam, ndraws = 1e3) %>%
#     data.frame() %>%
#     dplyr::select(participant, time, condition, .epred, .draw) %>%
#     group_by(participant, time) %>%
#     pivot_wider(names_from = condition, values_from = .epred) %>%
#     mutate(epred_diff = cond2 - cond1) %>%
#     summarise(m = mean(epred_diff > eps) ) %>%
#     mutate(prob_ratio = m / (1 - m) ) %>%
#     ungroup()

# plotting these predictions
# preds %>%
#     ggplot(aes(x = time, y = m) ) +
#     geom_line() +
#     facet_wrap(~participant)

# retrieving posterior samples
# posterior_samples <- posterior_epred(object = gp_model, newdata = time_seq)
# posterior_samples <- posterior_epred(object = gam, newdata = time_seq)
# library(modelr)
# meta_gam$data %>%
#     add_epred_draws(meta_gam, ndraws = 20, re_formula = NA) %>%
#     ggplot(
#         aes(
#             x = time, y = .epred,
#             colour = condition,
#             group = interaction(participant, condition, .draw)
#             )
#         ) +
#     geom_hline(yintercept = 0, linetype = 2) +
#     geom_vline(xintercept = true_onset, linetype = 2) +
#     geom_vline(xintercept = true_offset, linetype = 2) +
#     geom_line(show.legend = TRUE, alpha = 0.5, linewidth = 0.5) +
#     facet_wrap(~participant) +
#     labs(x = "Time (s)", y = "EEG signal (a.u.)")

# computing the difference between cond2 and cond1
# posterior_samples <- posterior_samples[, (ncol(posterior_samples)/2):ncol(posterior_samples)] -
#     posterior_samples[, 1:(ncol(posterior_samples)/2+1)]

# looping over different values of eps and threshold
for (i in seq_len(nrow(results) ) ) {

    # printing progress
    # cat("Assessing combination:", i, "out of", nrow(results), "combinations...")

    # retrieving current eps and threshold values
    eps <- results$eps[i]
    threshold <- results$threshold[i]

    # computing probability metrics with the current eps value
    # prob_y_above_0 <- data.frame(
    #     time = seq(min(ppt_df$time), max(ppt_df$time), length.out = n_approx*2-1)
    #     ) %>%
    # mutate(post_prob = colMeans(posterior_samples) ) %>%
    # mutate(m = colMeans(posterior_samples > eps) ) %>%
    # mutate(prob_ratio = m / (1 - m) )

>>>>

---
title: "Precise temporal localisation of M/EEG effects with Bayesian generalised additive multilevel models"
shorttitle: "Bayesian modelling of M/EEG data"
author:
  - name: Ladislas Nalborczyk
    corresponding: true
    orcid: 0000-0002-7419-9855
    email: ladislas.nalborczyk@cnrs.fr
    url: https://lnalborczyk.github.io
    affiliations:
      - id: id1
        name: "Aix Marseille Univ, CNRS, LPL"
        address: "5 avenue Pasteur"
        city: "13100 Aix-en-Provence"
        country: France
  - name: Paul Bürkner
    orcid: 0000-0001-5765-8995
    url: https://paulbuerkner.com
    affiliations:
      - name: "TU Dortmund University, Department of Statistics"
author-note:
  disclosures:
    conflict-of-interest: "The authors have no conflicts of interest to disclose."
    # financial-support: "This study was supported by Grant [Grant Number] from [Funding Source]."
abstract: "Time-resolved electrophysiological measurements such as those obtained through magneto- or electro-encephalography (M/EEG) offer a unique window into the neural activity underlying cognitive processes. Researchers are often interested in determining whether and when these signals differ across experimental conditions or participant groups. The conventional approach involves mass-univariate statistical testing across time, followed by corrections for multiple comparisons (e.g., FDR, FWER) or cluster-based inference. While effective for controlling error rates, cluster-based methods come with a significant limitation: they shift the focus of inference from individual time points to clusters, making it difficult to draw precise conclusions about the onset or offset of observed effects. Here, we introduce a *model-based* approch for analysing M/EEG timeseries such as event-related potentials (ERPs) or decoding performance over time. Our method leverages Bayesian generalised additive multilevel models (GAMMs), providing posterior probabilities that an effect is above zero (or above chance) at each time point, while naturally accounting for temporal dependencies and between-subject variability. Using both simulated and actual M/EEG datasets, we demonstrate that this approach substantially outperforms conventional methods in estimating the onset and offset of neural effects, yielding more precise and reliable results. We also provide an R package implementing the method and show how it can be seamlessly integrated into M/EEG analysis pipelines using MNE-Python."
keywords: [EEG, MEG, generalised additive models, mixed-effects models, multilevel models, Bayesian statistics, brms]
floatsintext: true
numbered-lines: true
bibliography:
  - meeg_modelling.bib
  - grateful-refs.bib
suppress-title-page: false
link-citations: true
# see https://www.overleaf.com/learn/latex/Font_typefaces
# and systemfonts::system_fonts()
# pdf-engine: xelatex
# monofont: Roboto
mainfont: CMU Serif
fontsize: 10pt
draftfirst: false
draftall: false
a4paper: true
donotrepeattitle: true
lang: en
toc: true
language:
  citation-last-author-separator: "&"
  email: "email"
  title-block-author-note: "Author note"
  title-block-correspondence-note: "Correspondence concerning this article should be addressed to"
  title-block-role-introduction: "Author roles were classified using the Contributor Role Taxonomy (CRediT; https://credit.niso.org/) as follows:"
format:
  # apaquarto-html: default
  apaquarto-pdf:
    latex-engine: xelatex
    documentmode: doc # man, jou, doc, stu
    keep-tex: true
    include-in-header:
      - text: |
          \usepackage{mathtools}
---

```{r setup, include = FALSE}
# loading packages
library(changepoint)
library(tidyverse)
library(tidybayes)
library(patchwork)
library(MetBrewer)
library(tidytext)
library(grateful)
library(ggrepel)
library(scales)
library(pakret)
library(scico)
library(knitr)
library(brms)

@@

# setting a default ggplot2 theme
# theme_set(theme_light(base_size = 10, base_family = "Open Sans") )
theme_set(theme_bw(base_size = 10, base_family = "Open Sans") )

@@

Understanding the temporal dynamics of cognitive processes requires methods that can capture fast-changing neural activity with high temporal resolution. Magnetoencephalography (MEG) and electroencephalography (EEG) are two such methods, widely used in cognitive neuroscience for their ability to track brain activity at the millisecond scale. These techniques provide rich time series data that reflect how neural responses unfold in response to stimuli or tasks. A central goal in many M/EEG studies is to determine whether—and when—neural responses differ across experimental conditions or participant groups.

@@

Description of cluster-based approaches [see @sassenhagen2019]...

However, as pointed by the original authors themselves [@maris2007], "there is a conflict between this interest in localized effects and our choice for a global null hypothesis: by controlling the FA \[false alarm\] rate under this global null hypothesis, one cannot quantify the uncertainty in the spatiotemporal localization of the effect" [@maris2007; @sassenhagen2019]. In contrast, the proposed approach, based on Bayesian generalised additive multilevel models, allows quantifying the posterior probability of effects being above chance at the level of timesteps, voxels, sensors (etc), while naturally taking into account spatiotemporal dependencies present in M/EEG timeseries.

@@

where $\beta_{jk}$ is the weight (coefficient) associated with the $k$th basis function $b_{jk}()$ evaluated at the covariate value $x_{ij}$ for the $j$th smooth function $f_{j}$. Splines coefficients are penalised (usually through the squared of the smooth functions' second derivative) in a way that can be interpreted, in Bayesian terms, as a prior on the "wiggliness" of the function. In other words, more complex (wiggly) basis function are penalised. To clarify the terminology at this stage: *splines* are functions composed of simpler functions. These simpler functions are basis functions (e.g., a cubic polynomial) and the set of basis functions is a *basis*. Each basis function gets its coefficient and the resultant spline is the sum of these weighted basis functions.

@@

Instead of averaging, obtain the smooth ERP signal from multilevel GAM... less susceptible to outliers [@meulman2023]...

@@

We then fitted a Bayesian GAM using the `brms` package [@brms2017; @brms2018; @nalborczyk2019]. We used the default priors in `brms` (i.e., weakly informative priors). We ran eight Markov Chain Monte-Carlo (MCMC) to approximate the posterior distribution, including each 5000 iterations and a warmup of 2000 iterations, yielding a total of $8 \times (5000-2000) = 24000$ posterior samples to use for inference. Posterior convergence was assessed examining trace plots as well as the Gelman–Rubin statistic $\hat{R}$. The `brms` package uses the same syntax as `r pkrt("mgcv")` for specifying smooth effects. @fig-plot-post-slope shows the predictions of this model together with the raw data.

@@

# timing the model fitting
# around 600s for 2000 iterations on 4 chains and 4 cores
# around 190s for 2000 iterations on 4 chains and 4 cores + cmdstanr and 01
# around 440s for 2000 iterations on 4 chains and 4 cores + threading(2), cmdstanr and 01
# library(tictoc)
# tic()

# fitting the GAM
gam <- brm(
    # cubic regression splines with k-1 basis functions
    # eeg ~ condition + s(time, bs = "cr", k = 20, by = condition),
    # thin-plate regression splines with k-1 basis functions
    eeg ~ condition + s(time, bs = "tp", k = 20, by = condition),
    data = ppt_df,
    family = gaussian(),
    warmup = 2000,
    iter = 5000,
    chains = 8,
    cores = 8,
    file = "models/gam.rds"
    #################################
    # speeding up the model fitting #
    #################################
    # backend = "cmdstanr",
    # threads = threading(2),
    # stan_model_args = list(stanc_options = list("O1") )
    )

# stop timing
# toc()
```

@@

@@

We then compute the posterior probability of the slope for `condition` being above $0+\epsilon$ (@fig-post-prob-ratio, left), with $\epsilon := 0.05$, which can be interpreted as the smallest effect size of interest. We can also express this as the ratio of posterior probabilities (i.e., $p/(1-p)$) and visualise the timecourse of this ratio superimposed with the conventional thresholds on evidence ratios (@fig-post-prob-ratio, right).

@@

# using the function with the two models
# plot_post_test(model = gam, eps = 0.1)
```

```{r tests, echo = FALSE, eval = FALSE}
# compute percentage in ROPE
library(easystats)
# rope(
#     x = gam,
#     range = "default",
#     ci = 0.95,
#     ci_method = "ETI",
#     effects = "fixed",
#     # component = c("conditional", "zi", "zero_inflated", "all"),
#     # parameters = NULL,
#     verbose = TRUE
#     )
# rope_range(x = gam)

# retrieving posterior draws for EEG over time
gam <- full_model
post_samples <- posterior_epred(object = gam, newdata = ppt_df, allow_new_levels = TRUE)

# defining the ROPE as baseline SD
# rope_range <- c(-0.1, 0.1)
baseline_sd <- ppt_df %>%
    pivot_wider(names_from = condition, values_from = eeg) %>%
    mutate(cond_diff = abs(cond1 - cond2) ) %>%
    filter(time < 0.1) %>%
    # head()
    summarise(baseline_sd = sd(cond_diff) ) %>%
    data.frame()

rope_range <- c(-baseline_sd, +baseline_sd)

# computing the HDI for each time step
hdi_values <- apply(post_samples, 2, function (x) hdi(x, ci = 0.95) )

# computing the  proportion of the HDI that falls within the ROPE
prop_in_rope <- sapply(1:ncol(post_samples), function (i) {
    lower_hdi <- hdi_values[[i]][1]
    upper_hdi <- hdi_values[[i]][2]
    # computing the  proportion inside ROPE
    mean(lower_hdi >= rope_range[1] & upper_hdi <= rope_range[2])
    })

# storing the results in a dataframe
rope_results <- data.frame(time = ppt_df$time, prop_in_rope = prop_in_rope)

# plotting the results
rope_results %>%
    summarise(prop_in_rope = mean(prop_in_rope), .by = time) %>%
    ggplot(aes(x = time, y = prop_in_rope) ) +
    geom_vline(xintercept = true_onset, linetype = 2) +
    geom_vline(xintercept = true_offset, linetype = 2) +
    geom_point()

# loading or fitting the full model
# full_model <- gam
full_model <- readRDS("models/gam.rds")

# fitting the null model (no 'condition' effect)
null_model <- brm(
    eeg ~ s(time, bs = "tp", k = 20),
    data = ppt_df,
    family = gaussian(),
    warmup = 2000,
    iter = 5000,
    chains = 8,
    cores = 8,
    file = "models/null_gam.rds"
    )

# computing the marginal likelihoods using bridge sampling
# library(bridgesampling)
# full_bridge <- bridge_sampler(full_model)
# null_bridge <- bridge_sampler(null_model)

# computing the Bayes Factor in favor of full model
# BF_10 <- bf(full_bridge, null_bridge)
# print(BF_10)

# extracting log-likelihoods for each model
log_lik_full <- log_lik(full_model)
log_lik_null <- log_lik(null_model)

# computing the pointwise log pointwise predictive density (LPD)
# averaging the log-likelihood across posterior draws
lpd_full <- colMeans(log_lik_full)
lpd_null <- colMeans(log_lik_null)

# computing the LPD difference (full model - null model)
lpd_diff <- lpd_full - lpd_null

# computing pointwise Bayes Factors as exp(lpd_diff)
BF_10_pointwise <- exp(lpd_diff)

# storing the results in a dataframe
ppt_df <- full_model$data
lpd_results <- data.frame(time = ppt_df$time, LPD_diff = lpd_diff, BF_10 = BF_10_pointwise)

# viewing the results
head(lpd_results)

# plotting it
lpd_results %>%
    # pivot_longer(cols = LPD_diff:BF_10) %>%
    summarise(BF_10 = mean(LPD_diff), .by = time) %>%
    # ggplot(aes(x = time, y = value, colour = name) ) +
    ggplot(aes(x = time, y = BF_10) ) +
    geom_vline(xintercept = true_onset, linetype = 2) +
    geom_vline(xintercept = true_offset, linetype = 2) +
    geom_line() +
    geom_point()
```

```{r fig-post-prob-ratio, echo = FALSE, out.width = "100%", fig.width = 9, fig.asp = 0.5, fig.cap = "Left: Posterior probability of the ERP difference (slope) being above 0 according to the GAM. Right: Ratio of posterior probability according to the GAM (on a log10 scale). Timesteps above threshold (10) are highlighted in green."}
# defining some parameters
n_approx <- 1e3
eps <- 0.05
threshold <- 10

@@

# computing the posterior probability that y is above 0
n_posterior_samples <- ndraws(gam)
prob_y_above_0 <- time_seq %>%
    add_epred_draws(object = gam) %>%
    data.frame() %>%
    dplyr::select(time, condition, .epred, .draw) %>%
    pivot_wider(names_from = condition, values_from = .epred) %>%
    mutate(epred_diff = cond2 - cond1) %>%
    group_by(time) %>%
    summarise(m = mean(epred_diff > 0 + eps) ) %>%
    ungroup() %>%
    mutate(prob_ratio = m / (1 - m) ) %>%
    # replacing infinite values by maximum possible values (number of posterior samples)
    mutate(prob_ratio = ifelse(is.infinite(prob_ratio), n_posterior_samples, prob_ratio) ) %>%
    mutate(prob_ratio = ifelse(prob_ratio == 0, 1/n_posterior_samples, prob_ratio) )

@@

# plotting it
gam_ratio_plot <- prob_y_above_0 %>%
    mutate(above_thres = ifelse(prob_ratio > threshold, 1, NA) ) %>%
    ggplot(aes(x = time, y = prob_ratio) ) +
    # geom_raster(
    #     data = bg_data,
    #     aes(x = time, y = prob_ratio, fill = fill_value),
    #     alpha = 0.5,
    #     interpolate = FALSE,
    #     inherit.aes = FALSE,
    #     show.legend = FALSE
    #     ) +
    # scale_fill_scico(palette = "vik", midpoint = log10(1) ) +
    geom_hline(yintercept = 1, linetype = "dashed") +
    geom_hline(yintercept = 3, linetype = "dashed", color = "darkred") +
    geom_hline(yintercept = 1/3, linetype = "dashed", color = "darkred") +
    geom_hline(yintercept = 10, linetype = "dashed", color = "red") +
    geom_hline(yintercept = 1/10, linetype = "dashed", color = "red") +
    geom_hline(yintercept = 100, linetype = "dashed", color = "orangered") +
    geom_hline(yintercept = 1/100, linetype = "dashed", color = "orangered") +
    geom_vline(xintercept = true_onset, linetype = 2) +
    geom_vline(xintercept = true_offset, linetype = 2) +
    geom_line(linewidth = 0.8) +
    geom_point(
        data = . %>% dplyr::filter(!is.na(above_thres) ),
        aes(y = threshold),
        colour = "darkgreen",
        shape = 16,
        size = 3,
        na.rm = TRUE
        ) +
    scale_y_log10(
        labels = label_log(digits = 2),
        breaks = c(1/1e4, 1/1e3, 1/1e2, 1/10, 1, 10, 1e2, 1e3, 1e4)
        # limits = c(0.0001, 1e4)
        ) +
    coord_cartesian(expand = FALSE) +
    labs(
        x = "Time (s)",
        y = expression(Pr(beta>0~"|"~data) / (1 - Pr(beta>0~"|"~data) ) )
        )

@@

# or plotting only the GAM
plot_post_test(model = gam, eps = 0.1) + gam_ratio_plot

# printing the results
# bind_rows(exceeding_times_gam, exceeding_times_gp) %>%
#     mutate(across(.cols = where(is.numeric), .fns = ~round(.x, 2) ) ) %>%
#     relocate(model, .before = cluster_onset)
```

@@

Next we fit a multilevel version of the GAM. Although it is possible to fit a GAMM at the single-trial level, we present a resource-efficient version of the model that is fitted directly on the by-participant summary statistics (mean and SD, similar to what is done in meta-analysis...

@@

We then computed the difference between the true and estimated onset/offset of the ERP difference ($\text{error}:=|\hat{\theta}-\theta|$), according to various `k` (GAM basis dimension) and `threshold` values. Remember that the signal is generated from a truncated Gaussian with an objective onset at 160 ms, a maximum at 250 ms, and an offset at 342 ms. @fig-onset-error shows that the hierarchical GAM can *exactly* recover the true onset and offset values, given some reasonable choice of `k` and `threshold` values (e.g., a threshold of 20). We provide more detailed recommendations on how to set `k` in @sec-basis.

```{r fig-onset-error, eval = TRUE, echo = FALSE, results = "hide", dpi = 300, fig.width = 9, fig.asp = 0.5, fig.cap = "Error function of onset (left) and offset (right) estimation according to various basis dimension and threshold values (according to the multilevel GAM)."}
# defining values of k and threshold to test
# threshold_values <- seq(from = 1, to = 50, by = 1)
# kvalues <- c(5, 10, 15, 20, 25, 30, 35, 40)

# number of simulations to run per combination
# nsims <- 10

# number of posterior samples to use
# n_post_samples <- 1e3

# importing the data generation functions 
# source(file = "code/generate_eeg_data.R")

# initialising results dataframe
# results <- crossing(kvalue = kvalues, threshold = threshold_values) %>%
# results <- crossing(kvalue = kvalues, sim_id = 1:nsims, threshold = threshold_values) %>%
#     mutate(
#         estimated_onset = NA,
#         estimated_offset = NA,
#         error_onset = NA,
#         error_offset = NA
#         )

# looping over different values of kvalue and threshold
# for (i in seq_len(nrow(results) ) ) {
# 
#     # retrieving current kvalue and threshold
#     kvalue <- results$kvalue[i]
#     threshold <- results$threshold[i]
# 
#     # printing progress
#     cat(
#         "\nAssessing combination:", i, "out of", nrow(results), "combinations...\n",
#         "k-value:", kvalue, "\n", "threshold:", threshold, "\n"
#         )
# 
#     # when we assess a new k-value
#     if (threshold == 1) {
#         
#         # generate some data
#         raw_df <- generate_data(
#             n_trials = n_trials, n_ppt = n_ppt, outvar = outvar,
#             srate = srate, ronset = ronset
#             )
#         
#         # sanity check
#         head(raw_df)
#         
#         # averaging across participants
#         summary_df <- raw_df %>%
#             summarise(
#                 eeg_mean = mean(eeg),
#                 eeg_sd = sd(eeg),
#                 .by = c(participant, condition, time)
#                 )
#         
#         # defining a contrast for condition
#         contrasts(summary_df$condition) <- c(-0.5, 0.5)
# 
#         # fitting the GAM
#         error_model <- brm(
#             # using by-participant SD of EEG across trials
#             eeg_mean | se(eeg_sd) ~
#                 condition + s(time, bs = "cr", k = kvalue, by = condition) +
#                 (1 | participant),
#             data = summary_df,
#             family = gaussian(),
#             warmup = 2000,
#             iter = 5000,
#             chains = 8,
#             cores = 8
#             )
# 
#     }
# 
#     prob_y_above_0 <- error_model$data %>%
#         add_epred_draws(object = error_model, ndraws = n_post_samples) %>%
#         # add_epred_draws(object = meta_gam) %>%
#         data.frame() %>%
#         dplyr::select(participant, time, condition, .epred, .draw) %>%
#         # dplyr::select(time, condition, .epred, .draw) %>%
#         pivot_wider(names_from = condition, values_from = .epred) %>%
#         mutate(epred_diff = cond2 - cond1) %>%
#         # computing mean posterior probability at the participant level
#         # group_by(participant, time) %>%
#         # summarise(m = mean(epred_diff > 0) ) %>%
#         # ungroup() %>%
#         # computing mean posterior probability at the group level
#         group_by(time) %>%
#         # summarise(m = mean(m) ) %>%
#         summarise(m = mean(epred_diff > 0) ) %>%
#         mutate(prob_ratio = m / (1 - m) ) %>%
#         ungroup()
# 
#     # finding onset, offset, and peak for the current threshold
#     exceeding_times <- prob_y_above_0 %>%
#         dplyr::filter(prob_ratio > threshold) %>%
#         summarise(
#             cluster_onset = min(time, na.rm = TRUE),
#             cluster_offset = max(time, na.rm = TRUE)
#             ) %>%
#         mutate(true_onset = true_onset, true_offset = true_offset)
# 
#     # storing the results in the dataframe
#     if (nrow(exceeding_times) > 0) {
# 
#         results$estimated_onset[i] <- exceeding_times$cluster_onset
#         results$estimated_offset[i] <- exceeding_times$cluster_offset
# 
#         # computing errors
#         results$error_onset[i] <- abs(
#             exceeding_times$cluster_onset - exceeding_times$true_onset
#             )
#         results$error_offset[i] <- abs(
#             exceeding_times$cluster_offset - exceeding_times$true_offset
#             )
# 
#     }
# 
# }

# saving the simulation results
# saveRDS(object = results, file = "results/meta_gam_error_properties.rds")
